---
title: "Wheat Trial Yields"
author: "Brittnany Graham"
format: html
editor: visual
---

### Part 1: Adding comments to the lines of code and what they do

Research Question:

Which combination of block and wheat variety produces the highest yield?

For this experiment, a block is a field that has different quality of soil which affects the total yield of each wheat variety different.
Within the block itself, there is one plot of each wheat variety, where the total plots is twenty.
The block structure matters because it takes into account the how different soil quality affects different wheat strains because some would do better in different types of soil because soil is not uniform across America.

```{r}
library(tidyverse)

set.seed(123) # setting the seed helps with reproducibility

# Define experimental structure
varieties <- c("A", "B", "C", "D", "E") # 5 wheat varieties
blocks <- c("Block1", "Block2", "Block3", "Block4") # the different field to account for variability, there are 5 fields in each block

# Define the truth
baseline_yield <- 3000  # kg/ha for variety A in block 1

variety_effects <- c(
  A = 0,
  B = 200,
  C = 350,
  D = 100,
  E = -50
) # adding what the true variety effects are to each of the varieties

block_effects <- c(
  Block1 = 0,
  Block2 = 150,
  Block3 = -100,
  Block4 = 200
) # block effect is the general effect of the field itself 

noise_sd <- 120

# Create all combinations of variety and block (full factorial design)
wheat_data <- expand.grid(
  variety = varieties,
  block = blocks
) # creating the data frame of all possibilities

# Add true yield based on variety and block effects
wheat_data <- wheat_data %>%
  mutate(
    variety_effect = variety_effects[variety],
    block_effect = block_effects[block],
    true_yield = baseline_yield + variety_effect + block_effect,
    yield = true_yield + rnorm(n(), mean = 0, sd = noise_sd)
  ) # updating the data frame with values indexed based on variety and block effects established above, then set an uninformative prior to not restrict the model and giving final yield

# Look at the data
head(wheat_data)
summary(wheat_data)

# Visualize
ggplot(wheat_data, aes(x = variety, y = yield, color = block)) +
  geom_point(size = 4, alpha = 0.7, position = position_dodge(width = 0.3)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, 
               color = "black", size = 0.3) +
  labs(
    title = "Simulated: Wheat Variety Yields Across Blocks",
    subtitle = "Each point = one plot; crossbar = variety mean across blocks",
    x = "Wheat Variety",
    y = "Yield (kg/ha)",
    color = "Block"
  ) +
  theme_minimal(base_size = 14)
```

### Part 2: Running models

#### Good model

```{r}
install.packages("lme4")
library(lme4)
install.packages("lmerTest")
library(lmerTest)

better_model <- lmer(yield ~ variety + (1|block), data = wheat_data)
summary(better_model)
anova(better_model)
```

There are a large difference between the residuals between the 'bad' model and the 'better' model because in the 'bad' model, the blocks are not taken into account, meaning the variation between fields and the results are therefore not reflected in the model, meaning the residuals are larger.
When the blocks are taken into account the residuals are smaller because it takes into account the noise and variation between the fields, which makes the residuals smaller and have more statistical power (larger F-statistic).
The means changed to reflect the variability and noise taken into account for each block.
The Intercept is the mean yield of variety A across all blocks.
In the subsequent variety categories, are how many units more or less than A they yielded.
So if variety A is 3187, then variety B had a mean of 3294 (3187+207).
I like to think of the intercept as the baseline, and the following units are what "shift" the baseline up or down, so you can see the yield of each variety.

#### What not to do

Simple anova model

```{r}

bad <- aov(yield ~ variety, wheat_data)  
summary(bad)
```

This models assumes that the blocks have the same effect or no effect on yield across all blocks.
If we ignore that block 2 has a higher crop yield, then we would overestimate the amount of which wheat does better generally.
Furthermore, the implication would mean that each field a wheat variety is grown in, would have to have the exact soil type as block 2 in this experiment.
If we ignore blocks, the variety estimates are inaccurate and useless to determine which wheat variety produces the highest yield.

**Visualizing it:**

```{r}
yield_plot <- ggplot(wheat_data, aes(variety, yield, color = block)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(title = "Yield by Variety and Block",
       x = "Variety",
       y = 'Yield',
       color = "Field Block")

yield_plot
```

### What I Learned:

The bad model is problematic because it does not include the blocking effects, more specifically, it doesn't take into account how the different variations of soil will affect the different wheat varieties.
Ignoring this block structure means the model will give inaccurate and unmeaningful results.
Even though we are looking at crop yield, we have to use a model that will take into account the block design and how each block *does* have an effect on yield because the soil quality is different.
This will glean more meaningful results because this agronomist understands that soil qualities across America is not uniform.
This will help farmers determine which wheat will produce the highest yield dependent on the quality of their soil.
